---
title: "043_panet_classification_summary"
author: "Marco Visani"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: true
---


# Introduction

Data received was of RGChannelSetExtended formal class. NOOB was run for both 450 and EPIC data. In order to include as much data as possible, low p value probes and low bead count probes were also included. Legacy probes were removed (total of 997 probes). The probes were sorted by probe name. Data was converted to beta values. Taking the intersect between the probe names of EPIC and 450K technology and the then data was joined together. Beta values were then normalized using `champ.norm`function. 

Normalized beta values were converted back to M values. Data was adjusted for batch effect on the `Slide` using ComBat. Values were then converted back to beta values. 

QC was then performed (see below).  

# Quality control 
Surrogate variables, PCA and t-SNE were checked. 

## Check for surrogate variables
First we checked for surrogate variables. None were found. 
```{r}
#rm(list = ls())
if(!exists("meta_data"))
  meta_data <- read.table("./data/meta_data/training_meta_data.txt", sep = "\t", header = T)
if(!exists("meth_combat_M"))
  meth_combat_M <- readRDS("./data/results/meth_combat_M.Rds")
if(!exists("meth_combat_beta"))
  meth_combat_beta <- as.data.frame(readRDS("./data/results/meth_combat_beta.Rds"))
if (!exists("meth_norm"))
  meth_norm <- as.data.frame(readRDS("./data/results/meth_normalized.Rds"))
```

```{r, fig.width = 14}
load(file = "./data/results/surrogate_var.RData")
require(ggplot2)
require(dplyr)
#this part will tell if the surrogate variables are descibing other varibales of the dataset
methSV <- as.data.frame(meth_sva$sv)
colnames(methSV) <- paste0("S", colnames(methSV))
methSV <- cbind(meta_data,methSV) 

SVList = as.list(rep(NA, times = sum(grepl("SV", colnames(methSV)))))
for (sv in seq_along(SVList)){
  SVList[[sv]] = lapply(colnames(meta_data), function(var){
    if(is.numeric(methSV[[var]]))
      test_res = summary(lm(as.formula(paste0("SV", sv,"~", var)), data = methSV))$coef[2,4]
    else
      test_res = kruskal.test(as.formula(paste0("SV", sv,"~", var)), data = methSV)$p.value
    data.frame(variable = var,
               pval = test_res,
               SV = paste0("SV", sv))
  })
}

#This part plot the 
bind_rows(SVList) %>% 
  mutate(pval_bin = cut(-log10(pval), 
                        breaks = c(0, -log10(0.05), 2, 5, 10, 300), 
                        labels = c("a", "b", "c", "d", "e"), 
                        include.lowest = T), SV = factor(SV, levels=paste0("SV", 1:42))) %>% 
  ggplot(aes(x = SV, y = variable, fill = pval_bin)) + 
  geom_tile() + 
  coord_fixed() + 
  scale_fill_manual(values = c("white", "pink", "orange", "red", "darkred"),
                    limits = c("a", "b", "c", "d", "e"),
                    labels = c("p >= 0.05", "p < 0.05", "p < 0.01", 
                               expression("p < 1x" ~ 10^{-5}), 
                               expression("p < 1x" ~ 10^{-10}))) +
  theme_classic() + 
  theme(legend.title = element_blank(), 
        legend.text.align = 0, 
        legend.background = element_rect(fill = "grey90"), 
        axis.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + 
  labs(title = "SVs vs biological variables")
```


## PCA
PCA tries to preserve the Global Structure of data i.e when converting d-dimensional data to d’-dimensional data then it tries to map all the clusters as a whole due to which local structures might get lost. That is why we also checked with t-SNE. 
```{r, fig.height=10}
require(fafreqs)
require(tibble)
#probevar <- rowVars(meth_combat_M)
probevar <- apply(meth_combat_M, 1, var)
probevar <- order(probevar, decreasing = T)

meth_combat_pca <- meth_combat_M[probevar[1:30000],meta_data$Sample_Name] #30000 is an arbitrary number --> look at paper on brain tumors (they took 32000)

meth_combat_pca <- t(meth_combat_pca)
pca_result <- prcomp(meth_combat_pca)

pca_df = pca_result$x %>% 
  as.data.frame() %>% 
  rownames_to_column("Sample_Name")
pca_df = left_join(pca_df, 
                   meta_data,
                   by = "Sample_Name")
pca_eig = factoextra::get_eigenvalue(pca_result)



cowplot::plot_grid(
  ggplot(pca_df, aes(color=Technology, x=PC1, y=PC2))+
    geom_point()+labs(x=paste0("PC1 ", pca_eig$variance.percent[1]), 
                       y=paste0("PC2 ", pca_eig$variance.percent[2])),
  
  ggplot(pca_df, aes(color=MEN1, x=PC1, y=PC2))+
    geom_point(),
  
  ggplot(pca_df, aes(color=DAXX_ATRX, x=PC1, y=PC2))+
    geom_point(),
  
  ggplot(pca_df, aes(color=Grade, x=PC1, y=PC2))+
    geom_point(),
  
  ggplot(pca_df, aes(color=MCT4_max, x=PC1, y=PC2))+
    geom_point(),
  
  ggplot(pca_df, aes(color=CC_Epi_newLRO, x=PC1, y=PC2))+
    geom_point(), ncol =2, align = "v" 
)

```
We can't see any pattern. Maybe a small one for the Intermidiate_ADM probes classified by the consensus cluster (bottom left image). 

## t-SNE
```{r, fig.height=10, fig.width=14}
require(Rtsne)
require(RColorBrewer)
set.seed(123) # setting a fixed seed allows reproducibility 
tsne <- Rtsne::Rtsne(meth_combat_pca, theta=0.0, PCA=F,
                     max_iter=2500) #with default perplexity of 30. Theta set to 0.0 for higher accuracy (but lower speed)
                                    # PCA and max_iter are set as in Capper et al. paper
tsne_df <- tsne$Y %>% as.data.frame() %>% rownames_to_column("Sample_Name")
tsne_df$Sample_Name <- c(rownames(meth_combat_pca))
tsne_df <- left_join(tsne_df, meta_data, by="Sample_Name")


cowplot::plot_grid(
  ggplot(tsne_df, aes(x=V1, y=V2, color=Technology))+
  xlab("tSNE1") + ylab("tSNE2")+
    geom_point(), 
  
  ggplot(tsne_df, aes(x=V1, y=V2, color=CC_Epi_newLRO))+
    xlab("tSNE1") + ylab("tSNE2")+
  geom_point(), 
  
  ggplot(tsne_df, aes(x=V1, y=V2, color=MEN1))+
    xlab("tSNE1") + ylab("tSNE2")+
  geom_point(), 
  
  ggplot(tsne_df, aes(x=V1, y=V2, color=DAXX_ATRX))+
    xlab("tSNE1") + ylab("tSNE2")+
  geom_point(), 
  
  ggplot(tsne_df, aes(x=V1, y=V2, color=Grade))+
    xlab("tSNE1") + ylab("tSNE2")+
  geom_point(), 
  
  ggplot(tsne_df, aes(x=V1, y=V2, color=MCT4_max))+
    xlab("tSNE1") + ylab("tSNE2")+
  geom_point(), ncol =2, align = "v"
)
```
Again there seems to be no significant cluster. Maybe Beta-like samples are a little bit isolated 

```{r quantification of nromalization effect}

getNormQual = function(norm_data,
                       raw_data){
  # this function is taken from the wateRmelon package (qual)
  # The na.rm = T settings are removed to have the function give an error if there are NAs
  # difference between normalized and raw
  stopifnot(all.equal(colnames(norm_data), colnames(raw_data)))
  dif  = norm_data - raw_data
  # root mean square deviation
  rmsd = sqrt(colMeans(dif^2))
  sdd  = apply(dif, 2, sd)
  sadd = apply(abs(dif), 2, sd)
  srms = rmsd/sdd
  data.frame(Sample_Name = colnames(norm_data),
             rmsd,
             sdd,
             sadd,
             srms)
}
 
```

 

```{r normalization violence, fig.height = 3.5, fig.width = 4.5}
norm_qual <- getNormQual(meth_norm, meth_combat_beta)
#detach("package:factoextra", unload=TRUE)
require(ggrepel)
require(ggplot2)
cowplot::plot_grid(
  left_join(meta_data,
            norm_qual,
            by = "Sample_Name") %>%
    ggplot(aes(x = rmsd,
               y = sdd,
               label = Sample_Name,
               color= Technology)) +
    geom_point(size = 3,
               alpha = 0.75) +
    geom_text_repel() +
    scale_color_brewer(palette = "Set1") +
    theme_classic() +
    labs(title = "NOOB + BMIQ")
)

```

# Classifier development
```{r}
rm(list = ls())
require(randomForest)

load("./results/rf.pred.RData")
if (!exists("meta_data"))
  meta_data <- read.table(file = "./data/meta_data/training_meta_data.txt", sep = "\t", header = T)
if (!exists("meth_combat_beta"))
  meth_combat_beta <- as.data.frame(readRDS("./data/results/meth_combat_beta.Rds"))

```

# Classifier cross-validation
Batch effect was NOT taken into account here !
```{r,echo=FALSE,message=FALSE,error=FALSE,warning=FALSE,fig.height=4,fig.width=12}
rm(list=ls())
library(pROC)
library(HandTill2001)
source(file.path("scripts", "R","multi_brier.R"))
source(file.path("scripts", "R","multi_logloss.R"))

load(file.path("results","CVresults.RData"))

AUCscores <- HandTill2001::auc(multcap(response=as.factor(y),predicted=scores))
AUCprobs <-  HandTill2001::auc(multcap(response=as.factor(y),predicted=probs))

errs <- sum(y!=ys)/length(y)
errp <- sum(y!=yp)/length(y)

briers <- brier(scores,y)
brierp <- brier(probs,y) 

logls <- mlogloss(scores,y)
loglp <- mlogloss(probs,y) 

out <- cbind(c(AUCscores,AUCprobs),c(errs,errp),c(briers,brierp),c(logls,loglp))
colnames(out) <- c("auc","misclassification","brier score","logloss")
rownames(out) <- c("raw scores","calibrated scores")

y.true <- y
y.score <- ys
y.calibrated <- yp
out
```

```{r}
table(y.true,y.score)
table(y.true,y.calibrated)
```

```{r, fig.height=10}
par(mfrow=c(3, 2))
for(i in 1:ncol(scores)){
  plot(x=scores[,i],y=probs[,i],ylim=c(0,1),pch=16,xlim=c(0,1),col=as.factor((y==colnames(scores)[i])),main=colnames(scores)[i] ,xlab="scores",ylab="calibrated scores")
  abline(a=0,b=1,col="green",lty=2)
}
```

```{r}
require(ggplot2)
require(ggridges)
raw_score <- as.data.frame(scores)
raw_score["predicted"] <- as.vector(y.score)
raw_score["CC_Epi_newLRO"] <- as.vector(y)
raw_score <- raw_score[raw_score$predicted == raw_score$CC_Epi_newLRO, ]

yl <- c()
for (i in 1:nrow(raw_score)) {
  x <- raw_score[i, 'predicted']
  yl <- append(yl, raw_score[i, x])
}
raw_score["max"] <- yl
rm(yl, x)

# plot the raw scores
raw <- ggplot(raw_score, aes(x = max, y = CC_Epi_newLRO, fill = CC_Epi_newLRO)) +
  ggtitle("Raw scores for correctly classified cases") +
  xlim(0,1.2) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

```{r}
require(ggplot2)
require(ggridges)
calibrated_score <- as.data.frame(probs)
calibrated_score["predicted"] <- as.vector(y.calibrated)
calibrated_score["CC_Epi_newLRO"] <- as.vector(y)
calibrated_score <- calibrated_score[calibrated_score$predicted == calibrated_score$CC_Epi_newLRO, ]

yl <- c()
for (i in 1:nrow(calibrated_score)) {
  x <- calibrated_score[i, 'predicted']
  yl <- append(yl, calibrated_score[i, x])
}
calibrated_score["max"] <- yl
rm(yl, x)

# plot the raw scores
calibrated <- ggplot(calibrated_score, aes(x = max, y = CC_Epi_newLRO, fill = CC_Epi_newLRO)) +
  xlim(0,1.2) +
  ggtitle("Calibrated scores for correctly classified cases") +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```



```{r, fig.width=10}
cowplot::plot_grid(
  raw, 
  calibrated,
  ncol = 1)

```